Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-31 paper.tex
First pass at Acks

Zarr is flexible in allowing different compression codecs and 
pre-compression filters to be specified on a per-array basis
(see section XXX).
Two key technologies often used in conjunction with Zarr are the Blosc
% This seems like the best citation for Blosc? It is mentioned here
meta-compressor~\cite{alted2010modern}
and Zstandard compression algorithm~\citep{collet2021rfc}.
Blosc is a high-performance compressor optimised for numerical
data which uses ``blocking''~\citep{alted2010modern} to 
optimise CPU-cache access patterns, as well as highly optimised
bit and byte shuffle filters.  Remarkably, on highly 
compressible datasets, Blosc decompression can be faster 
than \texttt{memcpy}.
% Mentioning this detail to reassure readers who might be thinking
% "how would this be implemented and what kind of horrible dependencies
% would I need?"
Blosc is written in C, with APIs for C, Python, Julia, Rust
and others.
Blosc is a ``meta-compressor'' because it provides 
access to several different compression codecs. The 
Zstandard codec is of particular 
interest here as it achieves very high compression ratios
with good decompression speeds. See Fig~\ref{fig-whole-matrix-decode}
for a comparison of Zstandard with LZ4 on decoding performance.
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is also used in several recent VCF compression 
methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
% LZ4 is also notable - any citations for this?

@@

\subsection{The VCF Zarr specification}
The VCF Zarr specification is a direct mapping from the VCF data model
to a chunked binary array format using Zarr,
and is evolution of the Zarr format used in the \texttt{scikit-allel}
package~\citep{miles2023scikit}.
VCF Zarr takes advantage
of Zarr's hierarchical structure by representing a VCF file as a top-level
Zarr group containing Zarr arrays. Each VCF field (fixed fields, INFO fields,
and FORMAT fields) is represented as a separate array in the Zarr hierarchy.
Some of the structures from the VCF header are also represented as arrays,
including contigs, filters, and samples.

@@

An important aspect of VCF Zarr is that field dimensions are 
global and fixed, and defined as the maximum across all rows.
Continuing the example above, the third dimension of the array 
is the maximum number of alternate alleles across \emph{all} 
variants. For variants at which there are less than the maximum
number of alternative alleles, the third dimension of the 
\texttt{call\_AD} array
is padded with a sentinel value (-2 for integers and a specific
non-signalling NaN for floats). While this is not a problem 
in practice for datasets in which all four bases are observed, 
it is a substantial issue for
fields that have a quadratic dependency on the number of alleles
(Number=G) such as PL. Such fields are already known to cause 
significant problems, and the ``local alleles'' proposal 
provides an elegant solution~\cite{poterba2024scalable}.
As this approach is on a likely path to 
standardisation~\cite{danecek2021twelve}, we would aim to 
include support in later versions of VCF Zarr.

@@

Once the VCF data has been converted to the intermediate columnar format,
it can then be converted to Zarr using the \texttt{vcf2zarr encode}
command. By default we choose integer widths based on the maximum
and minimum values observed during conversion to ICF
along with reasonable compressor defaults (see next section), 
but these can be
modified by generating a JSON-formatted storage schema, 
which can be edited and supplied as an argument to \texttt{encode}.
Encoding a given column (for example, \texttt{call\_AD})
involves creating a buffer to hold a full variant-chunk of the 
array in question, and then sequentially filling this buffer with 
values read from ICF and flushing to file.
 Similar to the \texttt{explode} command,
encoding to Zarr can be done in parallel on a single 
machine using the \texttt{encode} command,
or can be distributed across a cluster using the 
\texttt{dencode-init},
\texttt{dencode-partition} and \texttt{dencode-finalise} commands.
The distributed commands are fault-tolerant, reporting any failed
partitions so that they can be retried.

@@

\section{Data availability}

\section{Declarations}

@@

\begin{itemize}
    \item VCF: Variant Call Format;
    \item UKB: UK Biobank
    \item WGS: Whole Genome Sequence
\end{itemize}

\subsection{Competing Interests}
\subsection{Funding}

\subsection{Author's Contributions}

\section{Acknowledgements}

>>>>

Zarr is flexible in allowing different compression codecs and 
pre-compression filters to be specified on a per-array basis
(see section XXX).
Two key technologies often used in conjunction with Zarr are the Blosc
% This seems like the best citation for Blosc? It is mentioned here
meta-compressor~\cite{alted2010modern}
and Zstandard compression algorithm~\citep{collet2021rfc}.
Blosc is a high-performance compressor optimised for numerical
data which uses ``blocking''~\citep{alted2010modern} to 
optimise CPU-cache access patterns, as well as highly optimised
bit and byte shuffle filters.  Remarkably, on highly 
compressible datasets, Blosc decompression can be faster 
than \texttt{memcpy}.
% Mentioning this detail to reassure readers who might be thinking
% "how would this be implemented and what kind of horrible dependencies
% would I need?"
Blosc is written in C, with APIs for C, Python, Julia, Rust
and others.
Blosc is a ``meta-compressor'' because it provides 
access to several different compression codecs. The 
Zstandard codec is of particular 
interest here as it achieves very high compression ratios
with good decompression speeds (Figs~\ref{fig-whole-matrix-decode},
\ref{fig-compression-compressor}).
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is also used in several recent VCF compression 
methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
% LZ4 is also notable - any citations for this?

@@

\subsection{The VCF Zarr specification}
The VCF Zarr specification is a direct mapping from the VCF data model
to a chunked binary array format using Zarr,
and is an evolution of the Zarr format used in the \texttt{scikit-allel}
package~\citep{miles2023scikit}.
VCF Zarr takes advantage
of Zarr's hierarchical structure by representing a VCF file as a top-level
Zarr group containing Zarr arrays. Each VCF field (fixed fields, INFO fields,
and FORMAT fields) is represented as a separate array in the Zarr hierarchy.
Some of the structures from the VCF header are also represented as arrays,
including contigs, filters, and samples.

@@

An important aspect of VCF Zarr is that field dimensions are 
global and fixed, and defined as the maximum across all rows.
Continuing the example above, the third dimension of the array 
is the maximum number of alternate alleles across \emph{all} 
variants. For variants at which there are less than the maximum
number of alternative alleles, the third dimension of the 
\texttt{call\_AD} array
is padded with a sentinel value (-2 for integers and a specific
non-signalling NaN for floats). While this is not a problem 
in practice for datasets in which all four bases are observed, 
it is a substantial issue for
fields that have a quadratic dependency on the number of alleles
(Number=G) such as PL. Such fields are already known to cause 
significant problems, and the ``local alleles'' proposal 
provides an elegant solution~\cite{poterba2024scalable}.
As this approach is on a likely path to 
standardisation~\cite{danecek2021twelve}, we plan to 
include support in later versions of VCF Zarr.

@@

Once the VCF data has been converted to the intermediate columnar format,
it can then be converted to Zarr using the \texttt{vcf2zarr encode}
command. By default we choose integer widths based on the maximum
and minimum values observed during conversion to ICF
along with reasonable compressor defaults (see next section).
Default choices can be 
modified by generating a JSON-formatted storage schema,
which can be edited and supplied as an argument to \texttt{encode}.
Encoding a given field (for example, \texttt{call\_AD})
involves creating a buffer to hold a full variant-chunk of the 
array in question, and then sequentially filling this buffer with 
values read from ICF and flushing to file.
 Similar to the \texttt{explode} command,
encoding to Zarr can be done in parallel on a single 
machine using the \texttt{encode} command,
or can be distributed across a cluster using the 
\texttt{dencode-init},
\texttt{dencode-partition} and \texttt{dencode-finalise} commands.
The distributed commands are fault-tolerant, reporting any failed
partitions so that they can be retried.

@@

The VCF Zarr specification is available on GitHub at 
\url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
All source code for running benchmarks, analyses and creating 
"
plots in this article is available at
\url{https://github.com/sgkit-dev/vcf-zarr-publication}.
Vcf2zarr is freely available under the terms of the Apache 2.0 
license as part of the \texttt{bio2zarr} 
suite (\url{https://github.com/sgkit-dev/bio2zarr/}) 
and can be installed from the Python Package Index
(\url{https://pypi.org/project/bio2zarr/}).

@@

\begin{itemize}
    \item ICF: Intermediate Columnar Format
    \item GWAS: Genome Wide Association Study
    \item PBWT: Positional Burrows-Wheeler Transform
    \item QC: Quality Control
    \item UKB: UK Biobank
    \item VCF: Variant Call Format
    \item WGS: Whole Genome Sequence
\end{itemize}

% \subsection{Competing Interests}
\subsection{Funding}
JK acknowledges support from the Robertson Foundation, 
NIH (research grants HG011395 and HG012473), 
and Bill \& Melinda Gates Foundation (INV-001927; Sanger subaward G3594).
AM acknowledges support from the Bill \& Melinda Gates Foundation 
(INV-001927).

% \subsection{Author's Contributions}

\section{Acknowledgements}
This research was made possible through access to data in the National Genomic
Research Library, which is managed by Genomics England Limited (a wholly owned
company of the Department of Health and Social Care). The National Genomic
Research Library holds data provided by patients and collected by the NHS as
part of their care and data collected as part of their participation in
research. The National Genomic Research Library is funded by the National
Institute for Health Research and NHS England. The Wellcome Trust, Cancer
Research UK and the Medical Research Council have also funded research
infrastructure.

Computation used the Oxford Biomedical Research Computing (BMRC) facility, a
joint development between the Wellcome Centre for Human Genetics and the Big
Data Institute supported by Health Data Research UK and the NIHR Oxford
Biomedical Research Centre. The views expressed are those of the author(s) and
not necessarily those of the NHS, the NIHR or the Department of Health.
